---
title: "01-load-clean"
output: html_document
---

# Load & Clean Data

## 0. Library import

```{r}
library(readr)
library(tidyverse)
library(dplyr)
library(tidytext)
library(ggwordcloud)
library(visdat)
library(GGally)
library(viridis)
```

## 1\. Load Raw Data

```{r}
df <- read.csv('all_ai_models_final.csv', na.strings = c("", "NA"))
head(df, 10)
```

## 2\. NA visualization and percentages

```{r}
df_ordered <- df %>%
  select(order(colSums(is.na(.)), decreasing = FALSE))

vis_miss(df_ordered) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Missing Values Raw Data")

na_pct <- colMeans(is.na(df)) * 100
round(na_pct, 1)
```

## 3\. Drop columns with \>50% NA except Training.power.draw..W.

```{r}
cols_to_keep_after_na <- names(df)[
  (na_pct <= 50) | (names(df) == "Training.power.draw..W.")
]

df <- df[, cols_to_keep_after_na]

df_ordered <- df %>%
  select(order(colSums(is.na(.)), decreasing = FALSE))

vis_miss(df_ordered) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Missing Values after 50% NA filter")
```

## 4\. Drop rows where training power draw is NA

```{r}
df <- df %>% filter(!is.na(Training.power.draw..W.))

df_ordered <- df %>%
  select(order(colSums(is.na(.)), decreasing = FALSE))

vis_miss(df_ordered) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Missing Values after removing NA in power draw")
```

## 5\. Print NA percentages again & drop irrelevant columns

```{r}
na_pct_after <- colMeans(is.na(df)) * 100
round(na_pct_after, 1)

cols_to_drop <- c(
  "Training.code.accessibility",
  "Parameters.notes",
  "Training.compute.notes",
  "Dataset.size.notes",
  "Link",
  "Reference",
  "Last.modified",
  "Accessibility.notes",
  "Abstract"
)

df <- df %>% select(-all_of(cols_to_drop))

df_ordered <- df %>%
  select(order(colSums(is.na(.)), decreasing = FALSE))

vis_miss(df_ordered) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Heatmap of Missing Values")
```

## 6\. Create number of authors & impute missing values

```{r}
# number of authors
df <- df %>%
  mutate(num_authors = ifelse(
    is.na(Authors),
    NA,
    sapply(str_split(Authors, ","), length)
  ))

num_authors_mode <- names(sort(table(df$num_authors), decreasing = TRUE))[1] |> as.numeric()

df <- df %>%
  mutate(num_authors = ifelse(is.na(num_authors), num_authors_mode, num_authors))

df <- df %>% select(-Authors)

# impute categorical with mode
df <- df %>%
  mutate(across(where(is.character),
    ~ ifelse(is.na(.x),
      names(sort(table(.x), decreasing = TRUE))[1],
      .x)))

# impute numeric with median
df <- df %>%
  mutate(across(where(is.numeric),
    ~ ifelse(is.na(.x),
      median(.x, na.rm = TRUE),
      .x)))
```

## 7\. Clean duplicated comma-separated values

```{r}
df_clean <- df %>%
  mutate(across(c("Domain", "Task", "Organization", "Country..of.organization.", "Organization.categorization"),
    ~ str_split(.x, ",") %>%
      lapply(unique) %>%
      sapply(paste, collapse = ",")))
```

## 8\. Remove rows with remaining missing values

```{r}
df_clean <- df_clean %>% drop_na()
nrow(df_clean)
any(is.na(df_clean))
```

## 9\. Save cleaned dataset

```{r}
output_path <- "df_clean.csv"
write.csv(df_clean, output_path, row.names = FALSE)
output_path
```

