---
title: "06-stacked"
output: html_document
---
# Stacked Ensemble and Interpretability Analysis

## 1\. Library Import and Data Setup

```{r}
library(caret)
library(glmnet)
library(randomForest)
library(dplyr)
library(tidyverse)
library(patchwork)

set.seed(123) 
```

```{r}
V8 = read.csv("V8_log_scaled_onehot.csv")

target_name <- "training_power_draw_w"

V8_clean <- V8 %>% na.omit() 

train_idx <- createDataPartition(V8_clean[[target_name]], p = 0.8, list = FALSE)
train_data <- V8_clean[train_idx, ]
test_data <- V8_clean[-train_idx, ]

X_train <- train_data %>% select(-all_of(target_name))
y_train <- train_data[[target_name]]

X_test <- test_data %>% select(-all_of(target_name))
y_test <- test_data[[target_name]]

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:", nrow(test_data), "\n")
cat("Number of features:", ncol(X_train), "\n\n")
```

## 2\. Generate CV


```{r}
# Initialize prediction vectors
enet_preds <- numeric(nrow(train_data))
rf_preds   <- numeric(nrow(train_data))

# Create folds
folds <- createFolds(y_train, k = 5, list = TRUE)

cat("Starting 5-Fold Cross-Validation\n")

for(i in 1:5) {
  cat("\n=== Fold", i, "/5 ===\n")
  
  val_idx <- folds[[i]]
  train_idx_fold <- setdiff(1:nrow(train_data), val_idx)
  
  fold_X_train <- X_train[train_idx_fold, ]
  fold_y_train <- y_train[train_idx_fold]
  fold_X_val   <- X_train[val_idx, ]
  fold_y_val   <- y_train[val_idx]
  
  # --- Train Elastic Net ---
  enet_model <- cv.glmnet(
    x = as.matrix(fold_X_train),
    y = fold_y_train,
    alpha = 0.2,     # Elastic Net
    nfolds = 5
  )
  
  enet_preds[val_idx] <- as.vector(predict(
    enet_model,
    newx = as.matrix(fold_X_val),
    s = "lambda.min"
  ))
  
  # --- Train Random Forest ---
  rf_model <- randomForest(
    x = fold_X_train,
    y = fold_y_train,
    mtry = 9,
    ntree = 500
  )
  
  rf_preds[val_idx] <- predict(rf_model, newdata = fold_X_val)
  
  # RMSE for both models
  enet_rmse <- sqrt(mean((fold_y_val - enet_preds[val_idx])^2))
  rf_rmse   <- sqrt(mean((fold_y_val - rf_preds[val_idx])^2))
  
  cat("  Elastic Net RMSE:", round(enet_rmse, 3), "\n")
  cat("  RF RMSE:", round(rf_rmse, 3), "\n")
}

```

## 3\. Train Meta-Model


```{r}
cat("\n=== Training Meta-Model (Linear Regression) ===\n")

meta_train_data <- data.frame(
  target    = y_train,
  enet_pred = enet_preds,
  rf_pred   = rf_preds
)

meta_model <- lm(target ~ enet_pred + rf_pred, data = meta_train_data)
coefs <- coef(meta_model)

cat("\nMeta-Model Summary:\n")
print(summary(meta_model))

cat("\nMeta-Model Weights:\n")
cat("  Intercept:", round(coefs[1], 4), "\n")
cat("  Elastic Net weight:", round(coefs[2], 4), "\n")
cat("  RF weight:", round(coefs[3], 4), "\n")

meta_train_pred <- predict(meta_model, newdata = meta_train_data)
meta_train_rmse <- sqrt(mean((y_train - meta_train_pred)^2))

cat("\nMeta-model training RMSE (OOF):", round(meta_train_rmse, 3), "\n")

```
```{r}
weights <- data.frame(
  Model = c("Elastic Net", "Random Forest"),
  Weight = c(coefs[2], coefs[3])
)

library(ggplot2)
ggplot(weights, aes(x = Model, y = Weight, fill = Model)) +
  geom_bar(stat = "identity", width = 0.5) +
  geom_text(aes(label = round(Weight, 3)), vjust = -0.5) +
  ylim(0, 1) +
  labs(title = "Stacked Ensemble Meta-Model Weights",
       y = "Weight",
       x = "") +
  theme_minimal() +
  theme(legend.position = "none")

meta_train_data$pred <- meta_train_pred
ggplot(meta_train_data, aes(x = target, y = pred)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Stacked Ensemble Predictions vs Actual (Training OOF)",
       x = "Actual Target",
       y = "Predicted") +
  theme_minimal()

ggplot(meta_train_data, aes(x = enet_pred, y = rf_pred, color = pred)) +
  geom_point(alpha = 0.7) +
  labs(title = "Base Model Predictions and Stacked Ensemble Output",
       x = "Elastic Net Prediction",
       y = "Random Forest Prediction",
       color = "Final Ensemble") +
  theme_minimal()


```
## 4\. Train Final Base Models

```{r}
V8 = read.csv("V8_log_scaled_onehot.csv")

target_name <- "training_power_draw_w"

V8_clean <- V8 %>% na.omit() 

train_idx <- createDataPartition(V8_clean[[target_name]], p = 0.8, list = FALSE)
train_data <- V8_clean[train_idx, ]
test_data <- V8_clean[-train_idx, ]

X_train <- train_data %>% select(-all_of(target_name))
y_train <- train_data[[target_name]]

X_test <- test_data %>% select(-all_of(target_name))
y_test <- test_data[[target_name]]

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:", nrow(test_data), "\n")
cat("Number of features:", ncol(X_train), "\n\n")
cat("\n=== Training Final Models on Full Training Set ===\n")

# --- Final Elastic Net ---
final_enet <- cv.glmnet(
  x = as.matrix(X_train),
  y = y_train,
  alpha = 0.2,   # Elastic Net
  nfolds = 5
)

cat("Elastic Net optimal lambda:", round(final_enet$lambda.min, 6), "\n")

# --- Final Random Forest ---
final_rf <- randomForest(
  x = X_train,
  y = y_train,
  mtry = 9,
  ntree = 500
)

cat("Random Forest trained\n")

```
```{r}
# === Load V9 ===
V9 <- read.csv("V9_log_scaled_onehot_outlier.csv")

target_name <- "training_power_draw_w"

# Clean NA
V9_clean <- V9 %>% na.omit()

# Train/test split
train_idx_9 <- createDataPartition(V9_clean[[target_name]], p = 0.8, list = FALSE)
train_data_9 <- V9_clean[train_idx_9, ]
test_data_9  <- V9_clean[-train_idx_9, ]

# Features + target
X_train_9 <- train_data_9 %>% select(-all_of(target_name))
y_train_9 <- train_data_9[[target_name]]

X_test_9 <- test_data_9 %>% select(-all_of(target_name))
y_test_9 <- test_data_9[[target_name]]

cat("Training samples (V9):", nrow(train_data_9), "\n")
cat("Test samples (V9):", nrow(test_data_9), "\n")
cat("Number of features (V9):", ncol(X_train_9), "\n\n")
cat("\n=== Training Final Models on Full Training Set (V9) ===\n")

# --- Final Elastic Net (V9) ---
final_enet_9 <- cv.glmnet(
  x = as.matrix(X_train_9),
  y = y_train_9,
  alpha = 0,
  nfolds = 5
)

cat("Elastic Net optimal lambda (V9):", round(final_enet_9$lambda.min, 6), "\n")

# --- Final Random Forest (V9) ---
final_rf_9 <- randomForest(
  x = X_train_9,
  y = y_train_9,
  mtry = 9,
  ntree = 500
)

cat("Random Forest trained (V9)\n")
```

## 5\. Generate and Evaluate Test Predictions


```{r}

test_enet_pred_9 <- as.vector(predict(
  final_enet_9,         
  newx = as.matrix(X_test_9),
  s = "lambda.min"
))

test_rf_pred_9 <- predict(final_rf_9, newdata = X_test_9)

cat("\n=== Generating Test Set Predictions ===\n")

test_enet_pred <- as.vector(predict(
  final_enet,          
  newx = as.matrix(X_test),
  s = "lambda.min"
))

test_rf_pred <- predict(final_rf, newdata = X_test)

test_stack_data <- data.frame(
  enet_pred = test_enet_pred,
  rf_pred   = test_rf_pred
)

test_stack_pred <- predict(meta_model, newdata = test_stack_data)

cat("\n========================================\n")
cat("           FINAL TEST RESULTS\n")
cat("========================================\n\n")

rmse_enet_9  <- sqrt(mean((y_test_9 - test_enet_pred_9)^2))
rmse_rf_9    <- sqrt(mean((y_test_9 - test_rf_pred_9)^2))
rmse_stack <- sqrt(mean((y_test - test_stack_pred)^2))

rsq_enet_9  <- 1 - sum((y_test_9 - test_enet_pred_9)^2) / sum((y_test_9 - mean(y_test_9))^2)
rsq_rf_9    <- 1 - sum((y_test_9 - test_rf_pred_9)^2) / sum((y_test_9 - mean(y_test_9))^2)
rsq_stack <- 1 - sum((y_test - test_stack_pred)^2) / sum((y_test - mean(y_test))^2)

cat("Elastic Net V9:\n")
cat("  RMSE:", round(rmse_enet_9, 3), "\n")
cat("  R²:", round(rsq_enet_9, 3), "\n\n")

cat("RandomForest V9:\n")
cat("  RMSE:", round(rmse_rf_9, 3), "\n")
cat("  R²:", round(rsq_rf_9, 3), "\n\n")


cat("STACKED ENSEMBLE:\n")
cat("  RMSE:", round(rmse_stack, 3), "\n")
cat("  R²:", round(rsq_stack, 3), "\n\n")

best_single_rmse <- min(rmse_enet_9, rmse_rf_9)
improvement <- (best_single_rmse - rmse_stack) / best_single_rmse * 100

best_single_rsq <- max(rsq_enet_9, rsq_rf_9)
improvement_r2 <- (rsq_stack - best_single_rsq) / best_single_rsq * 100

cat("========================================\n")
cat("Improvement RMSE over best single model:", round(improvement, 2), "%\n")
cat("========================================\n")
cat("Improvement R2 over best single model:", round(improvement_r2, 2), "%\n")
cat("========================================\n")
```
```{r}
library(rpart)       
library(rpart.plot)  

tree_num <- 1
single_tree <- getTree(final_rf, k = tree_num, labelVar = TRUE)

print(single_tree)
single_rpart <- rpart(y_train ~ ., data = X_train, control = rpart.control(maxdepth = 5))
rpart.plot(single_rpart, type = 3, fallen.leaves = TRUE, main = "Single Decision Tree")
```

## 6\. Visualization and Diagnostics


```{r}
comparison_9 <- data.frame(
  Actual = y_test_9,
  ElasticNet_9 = test_enet_pred_9,
  RandomForest_9 = test_rf_pred_9
)

cat("\nCorrelation between V9 model predictions:\n")
cat("Elastic Net 9 vs RF 9:", round(cor(test_enet_pred_9, test_rf_pred_9), 3), "\n")

p1_9 <- ggplot(comparison_9, aes(x = Actual, y = ElasticNet_9)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Elastic Net (V9)", x = "Actual", y = "Predicted") +
  theme_minimal()

p2_9 <- ggplot(comparison_9, aes(x = Actual, y = RandomForest_9)) +
  geom_point(alpha = 0.5, color = "green") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Random Forest (V9)", x = "Actual", y = "Predicted") +
  theme_minimal()

combined_plot_9 <- p1_9 + p2_9
print(combined_plot_9)

comparison_long_9 <- comparison_9 %>%
  mutate(
    ElasticNet_Residual_9 = Actual - ElasticNet_9,
    RF_Residual_9 = Actual - RandomForest_9
  ) %>%
  select(Actual, ElasticNet_Residual_9, RF_Residual_9) %>%
  pivot_longer(
    cols = c(ElasticNet_Residual_9, RF_Residual_9),
    names_to = "Model", values_to = "Residual"
  )

ggplot(comparison_long_9, aes(x = Model, y = Residual, fill = Model)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Distribution by Model (V9)",
       y = "Residual (Actual - Predicted)") +
  theme_minimal() +
  theme(legend.position = "none")


comparison <- data.frame(
  Actual = y_test,
  ElasticNet = test_enet_pred,
  RandomForest = test_rf_pred,
  Stacked = test_stack_pred
)

cat("\nCorrelation between model predictions:\n")
cat("Elastic Net vs RF:", round(cor(test_enet_pred, test_rf_pred), 3), "\n")

p1 <- ggplot(comparison, aes(x = Actual, y = ElasticNet)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Elastic Net", x = "Actual", y = "Predicted") +
  theme_minimal()

p2 <- ggplot(comparison, aes(x = Actual, y = RandomForest)) +
  geom_point(alpha = 0.5, color = "green") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "RandomForest", x = "Actual", y = "Predicted") +
  theme_minimal()

p3 <- ggplot(comparison, aes(x = Actual, y = Stacked)) +
  geom_point(alpha = 0.5, color = "purple") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Stacked Ensemble", x = "Actual", y = "Predicted") +
  theme_minimal()

combined_plot <- p1 + p2 + p3
print(combined_plot)

comparison_long <- comparison %>%
  mutate(
    ElasticNet_Residual = Actual - ElasticNet,
    RF_Residual = Actual - RandomForest,
    Stack_Residual = Actual - Stacked
  ) %>%
  select(Actual, ElasticNet_Residual, RF_Residual, Stack_Residual) %>%
  pivot_longer(cols = c(ElasticNet_Residual, RF_Residual, Stack_Residual),
               names_to = "Model", values_to = "Residual")

ggplot(comparison_long, aes(x = Model, y = Residual, fill = Model)) +
  geom_boxplot() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "Residual Distribution by Model", y = "Residual (Actual - Predicted)") +
  theme_minimal() +
  theme(legend.position = "none")

```

## 7\. Interpretability and Diagnostics


### 7.1 Elastic Regression Coefficients (Feature Selection)


```{r}
cat("\n========================================\n")
cat("          INTERPRETABILITY ANALYSIS\n")
cat("========================================\n\n")

cat("=== ELASTIC NET: Feature Selection ===\n")

enet_coefs <- coef(final_enet, s = "lambda.min")
enet_coefs_df <- data.frame(
  Feature = rownames(enet_coefs),
  Coefficient = as.vector(enet_coefs)
) %>%
  filter(Feature != "(Intercept)") %>%
  mutate(
    Selected = Coefficient != 0,
    AbsCoef = abs(Coefficient)
  ) %>%
  arrange(desc(AbsCoef))

n_selected <- sum(enet_coefs_df$Selected)
n_total <- nrow(enet_coefs_df)

cat("\nFeature Selection Summary:\n")
cat("  Total features:", n_total, "\n")
cat("  Selected features:", n_selected, "\n")
cat("  Dropped features:", n_total - n_selected, "\n")
cat("  Feature reduction:", round((1 - n_selected/n_total) * 100, 1), "%\n\n")

cat("Top 15 Features by Absolute Coefficient:\n")
print(head(enet_coefs_df, 15))

write.csv(enet_coefs_df, "elastic_net_coefficients.csv", row.names = FALSE)

enet_plot_data <- enet_coefs_df %>%
  filter(Selected == TRUE) %>%
  arrange(desc(AbsCoef)) %>%
  head(20)

ggplot(enet_plot_data, aes(x = reorder(Feature, Coefficient), y = Coefficient)) +
  geom_col(aes(fill = Coefficient > 0)) +
  coord_flip() +
  labs(title = "Elastic Net: Top 20 Feature Coefficients", x = "Feature", y = "Coefficient") +
  theme_minimal() +
  scale_fill_manual(values = c("red", "blue"), 
                    labels = c("Negative", "Positive"),
                    name = "Effect")

```

### 7.2 SHAP Values for Random Forest (Local Explanations)


```{r}

library(fastshap)
library(randomForest)

set.seed(123)

n_shap <- min(500, nrow(X_test))
shap_idx <- sample(1:nrow(X_test), n_shap)
X_shap <- X_test[shap_idx, ]

pred_wrapper <- function(object, newdata) {
  predict(object, newdata = newdata)
}

shap_values <- fastshap::explain(
  object = final_rf,
  X = X_shap,
  pred_wrapper = pred_wrapper,
  nsim = 50,        
  adjust = TRUE
)

shap_importance <- data.frame(
  Feature = colnames(X_shap),
  Importance = colMeans(abs(shap_values))
) %>%
  arrange(desc(Importance))

print(head(shap_importance, 15))

```
```{r}
library(shapviz)

shap_viz <- shapviz(shap_values, X = X_shap)

sv_importance(shap_viz, kind = "beeswarm", max_display = 15)

top_feature <- shap_importance$Feature[1]
sv_dependence(shap_viz, v = top_feature)
```
## 8\. Stacked Model Contribution Analysis


```{r stacked_contribution}
cat("\n=== Stacked Model: Contribution Analysis ===\n")

stack_coefs <- coef(meta_model)

cat("Meta-Model Weights:\n")
cat("  Elastic Net contribution:", round(stack_coefs[2], 3), "\n")
cat("  RandomForest contribution:", round(stack_coefs[3], 3), "\n")

total_weight <- abs(stack_coefs[2]) + abs(stack_coefs[3])
enet_pct <- abs(stack_coefs[2]) / total_weight * 100
rf_pct   <- abs(stack_coefs[3]) / total_weight * 100

cat("\nPercentage Contribution:\n")
cat("  Elastic Net:", round(enet_pct, 1), "%\n")
cat("  RandomForest:", round(rf_pct, 1), "%\n\n")

test_contrib <- data.frame(
  Actual        = y_test,
  ENET_Pred     = test_enet_pred,
  RF_Pred       = test_rf_pred,
  ENET_Contrib  = test_enet_pred * stack_coefs[2],
  RF_Contrib    = test_rf_pred * stack_coefs[3],
  Intercept     = stack_coefs[1],
  Final_Pred    = test_stack_pred
)

cat("Example: First 5 predictions breakdown:\n")
print(head(test_contrib, 5))

write.csv(test_contrib, "stacked_model_contributions.csv", row.names = FALSE)

contrib_long <- test_contrib %>%
  select(ENET_Contrib, RF_Contrib) %>%
  pivot_longer(everything(), names_to = "Model", values_to = "Contribution")

ggplot(contrib_long, aes(x = Model, y = Contribution, fill = Model)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Model Contributions in Stacked Ensemble",
    y = "Contribution to Final Prediction"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```
